---
title: "Test Discrete Interim Analysis"
author: "Susannah Fleming"
date: "2023-11-16"
output: word_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```



```{r sourcing}
library(knitr)
source("DTAinterimAnalysis.R")
source("generateDTAdata.R")
# this provides some variables as well as functions.
```

```{r setupdata}

# load test data
testdata1 <- readRDS("testData1.rds")
testdata2 <- readRDS("testData2.rds")

# set up short test data
testshort1 <- testdata1[1:200,]
testshort2 <- testdata2[1:200,]
# these should nominally have Se= 65 and Sp = 85 but won't 

# calculate continuous Se and Sp
contshort1 <- continuousSeSp(testshort1)
contshort2 <- continuousSeSp(testshort2)

```

# Testing Interim analysis code

We have two short data sets (set 1 and set 2), both of which are n=200 from a larger data set of n=1000, and both of which have nominal Se of 65% and Sp of 85% and prevalence of 35% (exact values will be slightly different from this.)

The variables we can set are:

* *k*: number of interim analyses
* positions of interim analyses within the data set
* *alpha* - defaulting to 5% (alpha = 0.05)
* *p0* - the proportion defining H0: p <= p0
* p0 is separately defined for sensitivity and specificity


Planned testing:

* does the code perform as expected for different expected values of p0 and p1
* what is the effect of different numbers of interim analyses?
* is there actually a limit (5??) on the number of interim analyses? <- yes.  But you can't intuit it from the code
* (depending) can k be taken to the limit where we are effectively doing continuous interim analysis?

## Different values of p0 and p1

Test points at 20, 50 and 100.  Chosen for a useful variation of continuous Se and Sp.

```{r contsesp}
x <- 1:nrow(contshort1)
testpoints <- c(20, 50, 100)

# plot continuous Se and Sp for contshort1
plot(x,contshort1$Se,'l',col="red",ylim=c(0,1), main="Continuous Se and Sp, data set 1")
lines(x,contshort1$Sp,col="blue")
legend("bottomright", c("Sensitivity", "Specificity"), col=c("red","blue"),lty=c(1,1))
abline(v=testpoints)

# plot continuous Se and Sp for contshort2
plot(x,contshort2$Se,'l',col="red",ylim=c(0,1), main="Continuous Se and Sp, data set 2")
lines(x,contshort2$Sp,col="blue")
legend("bottomright", c("Sensitivity", "Specificity"), col=c("red","blue"),lty=c(1,1))
abline(v=testpoints)

# calculate TP, TN, allP and allN for each testpoint
TP <- c( sapply(testpoints, \(x) sum(contshort1$TP[1:x])),
         sapply(testpoints, \(x) sum(contshort2$TP[1:x])))
TN <- c( sapply(testpoints, \(x) sum(contshort1$TN[1:x])),
         sapply(testpoints, \(x) sum(contshort2$TN[1:x])))
allP <- c( sapply(testpoints, \(x) sum(contshort1$reference[1:x])),
         sapply(testpoints, \(x) sum(contshort2$reference[1:x])))
allN <- c( sapply(testpoints, \(x) sum(!contshort1$reference[1:x])),
         sapply(testpoints, \(x) sum(!contshort2$reference[1:x])))

# create table for output of actual sesp
sesptab <- data.frame(dataset=c(rep("Data set 1",3),rep("Data set 2",3)),
                      n=rep(testpoints,2),
                      Se = c(contshort1$Se[testpoints], contshort2$Se[testpoints]),
                      Sp= c(contshort1$Sp[testpoints], contshort2$Sp[testpoints]),
                      TP = TP,
                      TN = TN,
                      allP = allP, 
                      allN = allN)
kable(sesptab, digits=3)
                      

```

* Try correct, incorrect and variants around both

To test p0, we want to choose a variety that we expect to both trigger and not trigger at various points.
At this point, it doesn't matter whether we are looking at Se or Sp, so we can carry out independent tests on both.

Note that the Fleming method requires that the underlying analysis is being run on 1-Se and 1-Sp.

```{r testp0}

prevalence = 0.35
N = 200

# test changing p0 on Se
p0Se <- seq(0.5,0.9,0.1)
# test changing p1 on Sp
p0Sp <- seq(0.75, 0.95, 0.05)

res1ptests = list()
res2ptests = list()
op1ptests = data.frame()
op2ptests = data.frame()

for(i in 1:5) {
  # run each test
  suppressWarnings({
    test1 <- DTAdiscreteInterimAnalysis(contshort1,testpoints, pSe=p0Se[i], pSp=p0Sp[i], prevalence = prevalence, N=N)
    test2 <- DTAdiscreteInterimAnalysis(contshort2,testpoints, pSe=p0Se[i], pSp=p0Sp[i], prevalence = prevalence, N=N)
  })

  # collect results
  res1ptests[[i]] <- test1
  res2ptests[[i]] <- test2
  
  # convert to nice output
  op1ptests <- rbind(op1ptests, data.frame(Sensitivity= test1$Sensitivity$termstring,
                               Specificity= test1$Specificity$termstring))
  op2ptests <- rbind(op2ptests, data.frame(Sensitivity= test2$Sensitivity$termstring,
                               Specificity= test2$Specificity$termstring))
  

}

  
```

For Data Set1:

`r kable(op1ptests)`

For Data set 2:

`r kable(op2ptests)`


## Different values of k

p0 is set at: 
* 0.6 for sensitivity (expect stopping for efficacy as true Se is 65%)
* 0.9 for specificity (expect stopping for futility as true Sp is 85%)

Test points are evenly spaced for different values of k (e.g. k=3 will produce test points at 1/3 and 2/3 of the way through the dataset, and at the end.)

O'Brien and Fleming show why max k = 5, due to risks from multiple testing

```{r testk}
kvals <- c(2,3,4,5)

# function to generate evenly spaced test points
genEvenlySpacedTests <- function(k, n) {
  x <- round( (1:k) * n/k )
  return(x)
}

pSe <- 0.6 # stopping for efficacy
pSp <- 0.9 # stopping for futility

res1ktests = list()
res2ktests = list()
op1ktests = data.frame()
op2ktests = data.frame()
ktestdata1 = list()
ktestdata2 = list()


# run tests for each of kvals, with equally spaced points
for(i in 1:length(kvals)) {
  # run each test
  testpoints <- genEvenlySpacedTests(kvals[i], 200)
  suppressWarnings({
    test1 <- DTAdiscreteInterimAnalysis(contshort1,testpoints, pSe=pSe, pSp=pSp, prevalence = prevalence, N=N)
    test2 <- DTAdiscreteInterimAnalysis(contshort2,testpoints, pSe=pSe, pSp=pSp, prevalence = prevalence, N=N)
  })
  
  # collect results
  res1ktests[[i]] <- test1
  res2ktests[[i]] <- test2
  
  # collect test situation
  ktestdata1[[i]] <- data.frame(k=1:length(testpoints),
                                testpoints = testpoints,
                                Se=contshort1$Se[testpoints],
                                Sp=contshort1$Sp[testpoints])
  ktestdata2[[i]] <- data.frame(k=1:length(testpoints),
                                testpoints = testpoints,
                                Se=contshort2$Se[testpoints],
                                Sp=contshort2$Sp[testpoints])
  
  # convert to nice output
  op1ktests <- rbind(op1ktests, data.frame(k=kvals[i],
                                           Sensitivity= test1$Sensitivity$termstring,
                                           stopSe = contshort1$Se[testpoints[test1$Sensitivity$terminateAt]],
                                           stopatSe = testpoints[test1$Sensitivity$terminateAt],
                                           Specificity= test1$Specificity$termstring,
                                           stopSp = contshort1$Sp[testpoints[test1$Specificity$terminateAt]],
                                           stopatSp = testpoints[test1$Specificity$terminateAt]))
  op2ktests <- rbind(op2ktests, data.frame(k=kvals[i],
                                           Sensitivity= test2$Sensitivity$termstring,
                                           stopSe = contshort2$Se[testpoints[test2$Sensitivity$terminateAt]],
                                           stopatSe = testpoints[test2$Sensitivity$terminateAt],
                                           Specificity= test2$Specificity$termstring,
                                           stopSp = contshort2$Sp[testpoints[test2$Specificity$terminateAt]],
                                           stopatSp = testpoints[test2$Specificity$terminateAt]))
  
  
}


```

In the tables below, "stopSe" is the sensitivity at which the termination threshold is met, and "stopatSe" is the corresponding data point (noting that the samples are 200 points long, so termination at point 200 is equivalent to no early termination.)  Specificity is handled similarly.

The test with k=200 is equivalent to continuous interim analysis, as there is a test point at every data point.

For Data Set1:

`r kable(op1ktests,digits=3)`

at k=5:

`r kable(ktestdata1[[which(kvals==5)]])`


For Data set 2:

`r kable(op2ktests,digits=3)`

at k=5:

`r kable(ktestdata2[[which(kvals==5)]])`


The following graphs illustrate where the method terminates.

```{r whereterminate}
# illustrate where the method terminates

plotSeSpFlemingThresholds(contshort1, 0.6, 0.9)
plotSeSpFlemingThresholds(contshort2, 0.6, 0.9)

```

Manually confirmed that crossing points are where termination first occurs.

Note that, since the termination lines depend on p0, rather than on the data, they are identical for both datasets for a given p0 (i.e. sensitivity or specificity).  The "stepping" behaviour seen early in the studies is probably an artefact of the requirement for stopping values to be whole numbers (of events), and hence rounding effects when this is converted to a proportion for graphical illustration.

The graphs converge at a value below p0, but the H0 limit appears to cross the nominal p0 line at approximately n/2.

Test convergence of various alphas with p0=0.7 and ns from 200 to 10,000.

```{r checkalphaconvergence}
# model termination thresholds for different values of alpha

alphas <- c(0.01, 0.02, 0.05, 0.1)
p0s <- 0.7
ns <- c(200,500,1000,10000)

nalphares <- data.frame()
# it looks like the values are actually converging on p0, depending on n and alpha
for(a in alphas) {
  for(nk in ns){
    termthresh = modelFlemingTerminationThresholds(p0=1-p0s, n=nk, alpha=a)
    testres <- data.frame(p0=p0s[1],
                          alpha=a,
                          n=nk,
                          H0lim=termthresh$invH0limit[nk],
                          H1lim=termthresh$invH1limit[nk],
                          H0at50pc=termthresh$invH0limit[nk/2])
    nalphares <- rbind(nalphares,testres)
  }
}


```
`r kable(nalphares)`

It appears that the two limits converge closer to p0 as both alpha and n increase, and that the H0 limit at n/2 is approximately p0.  It's not clear why this is the case.


Next steps:

* try some other methods (Stafford/todd and/or lan/demetz), particularly to see if they perform similarly in continuous configuration
* a plot in ROC space of how sens and spec change over time?
i.e. a standard ROC plot of sens against spec, with a series of points or lines showing a two-dimensional 'random walk' of the sensitivity and specificity estimates as they evolve.
The 'rejection region' would then look like gradually shrinking rectangle (at least assuming sens and spec are being treated independently - otherwise it might be an ellipse) - this might be difficult to plot on a static graph, as it also changes over time.
The whole graph might end up looking too messy, but worth a try? I think it could help to emphasise one of the discussion points we want to make, that in diagnostic studies the decision would be based on the sens and spec together, rather than just one of them.





```{r createsavedgraphs, eval=FALSE}
# This code DOES NOT run when the file is knit.  It should be run manually
# when saved graphs are to be created.
library(ggplot2)
library(tidyverse)
library(ggrepel)

#### graphs of example data

# add x data to contshorts
contshort1$x = 1:nrow(contshort1)
contshort2$x = 1:nrow(contshort2)

# reshape for plotting
cs1 <- contshort1 %>% gather("Sesp", "Sespvalue", Se, Sp)
cs1$dataset <- "Dataset 1"
cs2 <- contshort2 %>% gather("Sesp", "Sespvalue", Se, Sp)
cs2$dataset <- "Dataset 2"
csboth <- merge(cs1,cs2, all=T)

# improve labelling
csboth$Sesp[csboth$Sesp=="Se"] <- "Sensitivity"
csboth$Sesp[csboth$Sesp=="Sp"] <- "Specificity"

csplot <- ggplot(data=csboth, aes(x=x, y=Sespvalue, colour=Sesp, label=Sesp)) + 
  geom_line() +
  theme_bw() +
  scale_colour_grey(name="Measure") +
  labs(x="Data point", y="Proportion") +
  theme(legend.position="right",
        text=element_text(size=16)) +
  facet_grid(. ~ dataset)
print(csplot)

ggsave("datasets.png",width=12,height=5)


### graphs of continuous thresholds
p0Se <- 0.6
p0Sp <- 0.9

# do some of the manual data creation
dataSe1 <- contshort1[contshort1$reference==T, ]
NSe1 <- nrow(dataSe1)
Sethresh1 <- modelFlemingTerminationThresholds(1-p0Se, n=NSe1)

dataSe2 <- contshort2[contshort2$reference==T, ]
NSe2 <- nrow(dataSe2)
Sethresh2 <- modelFlemingTerminationThresholds(1-p0Se, n=NSe2)

# specificity data, ref = F
dataSp1 <- contshort1[contshort1$reference==F, ]
NSp1 <- nrow(dataSp1)
Spthresh1 <- modelFlemingTerminationThresholds(1-p0Sp, n=NSp1)
dataSp2 <- contshort2[contshort2$reference==F, ]
NSp2 <- nrow(dataSp2)
Spthresh2 <- modelFlemingTerminationThresholds(1-p0Sp, n=NSp2)

# add xs and identifying data
Sethresh1$p0 <- Sethresh2$p0 <- p0Se
Spthresh1$p0 <- Spthresh2$p0 <- p0Sp
dataSe1$x <- Sethresh1$x <- 1:NSe1
dataSe2$x <- Sethresh2$x <- 1:NSe2
dataSp1$x <- Spthresh1$x <- 1:NSp1
dataSp2$x <- Spthresh2$x <- 1:NSp2
dataSe1$Sesp <- dataSe2$Sesp <- Sethresh1$Sesp <- Sethresh2$Sesp <- "Sensitivity"
dataSp1$Sesp <- dataSp2$Sesp <- Spthresh1$Sesp <- Spthresh2$Sesp <- "Specificity"
dataSe1$dataset <- dataSp1$dataset <- Sethresh1$dataset <- Spthresh1$dataset <- "Dataset 1"
dataSe2$dataset <- dataSp2$dataset <- Sethresh2$dataset <- Spthresh2$dataset <- "Dataset 2"

# gather and merge thresholds
Sethresh1 <- Sethresh1 %>% pivot_longer(cols=rg:p0, names_to="Measure", values_to="threshval")
Sethresh2 <- Sethresh2 %>% pivot_longer(cols=rg:p0, names_to="Measure", values_to="threshval")
Spthresh1 <- Spthresh1 %>% pivot_longer(cols=rg:p0, names_to="Measure", values_to="threshval")
Spthresh2 <- Spthresh2 %>% pivot_longer(cols=rg:p0, names_to="Measure", values_to="threshval")
allthresh <- rbind(Sethresh1, Sethresh2, Spthresh1, Spthresh2)

# improve labelling
allthresh$Measure[allthresh$Measure=="invH0limit"] <- "efficacy"
allthresh$Measure[allthresh$Measure=="invH1limit"] <- "futility"
# remove unneeded data 
allthresh <- allthresh %>% filter(Measure %in% c("efficacy", "futility", "p0"))

# tidy and merge data
dataSe1$Sespvalue <- dataSe1$Se
dataSp1$Sespvalue <- dataSp1$Sp
dataSe2$Sespvalue <- dataSe2$Se
dataSp2$Sespvalue <- dataSp2$Sp
keepcols <- c("x", "Sesp", "dataset", "Sespvalue")
dataSe1 <- dataSe1[ , keepcols ]
dataSp1 <- dataSp1[ , keepcols ]
dataSe2 <- dataSe2[ , keepcols ]
dataSp2 <- dataSp2[ , keepcols ]
alldata <- rbind(dataSe1, dataSe2, dataSp1, dataSp2)

plotdata <- merge(allthresh, alldata, by=c("x", "Sesp", "dataset"))


legendlabels= c("data", "p0", "efficacy", "futility")

threshplot <- ggplot(data=plotdata, aes(x=x, y=threshval, color=Measure, linetype=Measure)) + 
  geom_line(aes(y=Sespvalue, color="data", linetype="data")) +
  geom_line() +
  theme_bw() +
  scale_color_manual(name="Measure",
                     labels=legendlabels,
                     breaks=legendlabels,
                     values=c("grey70","black","black","black")) +
  scale_linetype_manual(name="Measure",
                        labels=legendlabels,
                        breaks=legendlabels,
                        values=c("solid", "solid", "longdash","dashed")) +
  theme(legend.position="right",
        text=element_text(size=16)) +
  labs(x="Data point", y="Proportion") +
  facet_grid(Sesp ~ dataset)
print(threshplot)

ggsave("thresholds.png",width=12,height=8)


```



